---
title: "Practical 5"
author: "James Hollway"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical 5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setting up

For this lab, we'll use a few different packages for data and analysis.
We're going to use the `m182` dataset from last week, but directly from the package this week.

```{r setup}
# library(igraph)
library(migraph) # note that you may need a special version for what follows...
library(ggplot2)
data("ison_m182", package = "migraph")
```

The network is anonymous, but I think it would be nice to add some names,
even if it's just pretend. Luckily, I've added a function for this.
This makes plotting the network just a wee bit more accessible:

```{r addingnames}
ison_m182 <- to_named(ison_m182)
autographr(ison_m182)
```

There are actually three different types of tie here. 
Let's separate them out into separate networks.

```{r separatingnets}
(m182_friend <- to_simplex(ison_m182, "friend_tie"))
gfriend <- autographr(m182_friend) + ggtitle("Friendship")
(m182_social <- to_simplex(ison_m182, "social_tie"))
gsocial <- autographr(m182_social) + ggtitle("Social")
(m182_task <- to_simplex(ison_m182, "task_tie"))
gtask <- autographr(m182_task) + ggtitle("Task")
grid.arrange(gfriend, gsocial, gtask, ncol = 3)
```

