---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=F}
library(migraph)
```

This small vignette will guide you through a step-by-step setup guide of some of the
tools you need to get started in network analysis. Note that this vignette does
not aim at being *the* comprehensive guide to network analysis but rather
strives to provide the reader with a lay of the land of the field of Social
Network Analysis. Check out the [other helpful resources section](#otherresources)
below if you'd like to go further.

## Before getting started; Installing R and R-Studio:

While there exist other tools to perform network analysis such as [Python](https://valdanchev.github.io/reproducible-data-science-python/network_analysis.html#open-network-data-repositories),
the majority of the tools for social network analysis are written in R. We will
therefore use the latter for this introduction to the analysis of social
networks.

Since this vignette would be too long and that there is a myriad of great
resources that explain in detail how to setup R and RStudio, we will assume
that the reader has already performed these steps.

- [A great guide to install R and RStudio by Jenny Brian](https://stat545.com/install.html)
- [Another guide to install R and RStudio](https://coursedown.netlify.app/labs/00-install_name_plot.html)

For the curious reader, here is a guide to setting up version control (read
Google Drive for programming) alongside your R installation. Note that this
step is optional for what will follow but will help prevent inadvertent loss
of your projects and will help you showcase them.

- [Happy Git with R by Jenny Brian](https://happygitwithr.com/index.html)

## Actually getting started; SNA packages:

As you have read in the guides above, R's great strength lies in its large
ecosystem of packages. They contain functions that enable you to wrangle,
analyse, and visualize data namely in the field of social network analysis.
The present section will link to a few of commonly used ones.

| Package   | Description                                                                                   | Link                                                       |
|-----------|-----------------------------------------------------------------------------------------------|------------------------------------------------------------|
| `{migraph}`   | migraph contains a comprehensive toolset for importing, handling, and analysing networks | https://snlab-ch.github.io/migraph/index.html              |
| `{igraph}`    | Popular network analysis toolset                                                              | http://igraph.org/r/                                       |
| `{tidygraph}` | Popular tidy network analysis toolset                                                    | https://www.data-imaginist.com/2017/introducing-tidygraph/ |
| `{ggraph}`    | ggplot style network vizualisation tools. Complement to tidygraph.                       | https://ggraph.data-imaginist.com/                         |
| `{statnet}`   | Integrated set of tools for Social Network Analysis.                                     | http://statnet.org/                                        |

Install them and explore their tools by reading the documentation. E.g. run
the following chunk to install `{migraph}` and explore its documentation.

```{r, eval=FALSE}
# Install migraph from CRAN
install.packages("migraph")

# Uncomment to install the latest stable version from GitHub directly
# install.packages("devtools")
# devtools::install_github("snlab-ch/migraph")

# Load migraph and explore its documentation
library(migraph)
help(package = "migraph")
```

Finally, there is a multitude of network analysis packages for R. This implies
that functions across packages are likely to be named in a similar way. To
prevent unexpected behavior in your scripts when loading multiple SNA packages,
we recommend that you use the following syntax to specify the *exact* function
you intend to use.

```{r, eval=FALSE}
# Use
migraph::as_igraph()
# Instead of
library(migraph)
as_igraph()
```

## We have the tools now; what about the data?

This list will get you started in your search for existing network data.

| Source                     | Description                                                                                                                                       | Link                                                                |
|----------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------|
| R   packages               | Packages often contain datasets   that<br />are useful for illustrative purposes.                                                                 | E.g.   https://snlab-ch.github.io/migraph/reference/index.html#data |
| Awesome   Network Analysis | Awesome-style list with a   dataset section                                                                                                       | https://github.com/briatte/awesome-network-analysis#datasets        |
| networkdata                | R package wit 979 network   datasets<br />containing 2135 networks                                                                                | https://github.com/schochastics/networkdata                         |
| Kateto   Blog              | Prof. Ognyanova's blog contains   a small<br />list of dataset sources that are great to start your first<br />Social network analysis project. | https://kateto.net/2016/05/network-datasets/                        |                                           |

Exploring the datasets present in the installed packages can be done within
R by typing the following:

```{r, eval=FALSE}
# Checking out the datasets in {migraph}
data(package = "migraph")

# Importing the dataset into your environment
your_data <- migraph::ison_marvel_relationships
```

## Other helpful resources in social network analysis: {#otherresources}

Want to know more about network analysis tools or data sources?
Then your first step should be to check out the *extremely comprehensive*
[Awesome Network Analysis page](https://github.com/briatte/awesome-network-analysis#datasets)
on GitHub. This curated list also provides its reader with books covering
the theory of SNA as well as a selection of papers applying SNA to a variety
of social sciences.

The [Kateto Blog](https://kateto.net) is equally as
great of a resource to get started and understand the variety of tools available
to social scientists analysing networks.

Finally note that this note focuses on the main tools to *handle* network data
and not so much on more advanced inferential tools. The latter are covered in
the other "Practical" vignettes here on the `{migraph}` website.
