---
title: "Practical 7"
author: "James Hollway"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Practical 7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setting up

For this lab, we'll explore a couple of different datasets.
First, let's examine heterogeneity within the Marvel relationships dataset.
Since this is a signed network, let's first extract just the friendships.
We'll also concentrate on just the main component (no isolates).
And, for the sake of simplicity, just the characters that appear
in the comics more than average for these characters.

```{r setup}
library(migraph)
marvel_friends <- to_unsigned(ison_marvel_relationships, keep = "positive")
marvel_friends <- to_main_component(marvel_friends)
marvel_friends <- marvel_friends %>% filter(Appearances >= mean(Appearances))
marvel_friends
```

This gives us a dataset of nearly twenty characters and a little more than 100 edges.
Recall that this data has several nodal attributes.
Let's explore a couple of these attributes visually.

```{r vis}
autographr(marvel_friends, 
           node_shape = "Gender",
           node_color = "PowerOrigin")
```

There seems to be a bit of a distribution of some of these variables.

# Calculating diversity

## Calculating Blau index

We can begin by calculating the Blau diversity for each attribute.

```{r blau}
graph_blau_index(marvel_friends, "Gender")
graph_blau_index(marvel_friends, "PowerOrigin")
graph_blau_index(marvel_friends, "Attractive")
graph_blau_index(marvel_friends, "Rich")
graph_blau_index(marvel_friends, "Intellect")
```

Looks like there is more diversity in terms of where these characters got
their powers, whether they have significant intellectual powers,
and their gender, than their attractiveness or their wealth.

We can also cross-reference this diversity.
For example, we might be interested in whether our comic book heroes
are equally gender diverse across their (power) origin stories,
or equally intellectually diverse across gender.^[Note that this works for calculated categorical variables too, such as cluster/group assignment from community detection or equivalence classes.]

```{r crossref}
graph_blau_index(marvel_friends, "Gender", "PowerOrigin")
graph_blau_index(marvel_friends, "Intellect", "Gender")
```

Looks like some origin stories are much more diverse than others.
Gods (just Thor here) and humans are all men,
whereas those with mutant or radiation origin stories are more gender diverse.
There doesn't seem to be any significant difference in intellect 
across gender categories however.
We can visualise this quite effectively using the `node_groups` argument:

```{r blaugroups}
autographr(marvel_friends, 
           node_group = "PowerOrigin", 
           node_shape = "Gender")
autographr(marvel_friends, 
           node_group = "Gender", 
           node_shape = "Intellect")
```

## Calculating EI index

Calculating the EI index follows the same syntax.
Remember that a negative EI index indicates that there are more
ties between nodes of the same class than between nodes of different classes.

```{r ei}
(obs.gender <- graph_ei_index(marvel_friends, "Gender"))
(obs.powers <- graph_ei_index(marvel_friends, "PowerOrigin")) 
(obs.attract <- graph_ei_index(marvel_friends, "Attractive")) 
```

Looks like there might be some gender homophily present,
but the score for power origin homophily is so close to 0 that
it does not seem to signal much.
There seems to be a fairly large effect for homophily on the basis of looks though...

But ultimately these are just scores,
and this doesn't tell us whether this is any more or less than
what we might expect the score to be by chance for a network
of this size and density and distribution of that attribute.

To get at that, we will simulate a series of random graphs
(Erdös-Renyi/Bernoulli) of the same dimensions and
distribution of the attribute to find out whether there is
more homophily or heterophily than expected by chance.

```{r rando}
(rand.gender <- test_random(marvel_friends, graph_ei_index, attribute = "Gender"))
(rand.power <- test_random(marvel_friends, graph_ei_index, attribute = "PowerOrigin"))
(rand.attract <- test_random(marvel_friends, graph_ei_index, attribute = "Attractive"))
par(mfrow=c(1,3))
plot(rand.gender)
plot(rand.power)
plot(rand.attract)
```

This is really interesting.
It looks like we cannot reject the null hypothesis that there is no homophily
for gender nor for attractiveness,
but we can reject the null hypothesis with respect to their power origin story.
While the effect itself is close to 0 (neither strong homophily nor heterophily),
all the random networks generated returned larger EI scores, between .1 and .4.
So there is significantly less heterophily here than expected.
