---
title: "Visualisation"
author: "by James Hollway"
output: 
  learnr::tutorial:
    theme: journal
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(manynet)
knitr::opts_chunk$set(echo = FALSE)
```

## Why we graph

Network visualisation is important and non-trivial.
As Tufte (1983: 9) said:

> "At their best, graphics are instruments for reasoning about quantitative information. Often the most effective way to describe, explore, and summarize a set of numbers – even a very large set – is to look at pictures of those numbers"

Brandes et al (1999) argue that visualising networks demands thinking about:

- _substance_: a concise and precise delivery of insights to the researcher and/or readers
- _design_: the ergonomics of function are 98% of the purpose of good design, aesthetics only 2%
- and _algorithm_: the features of the e.g. the layout algorithm

While there may be many dead-ends to exploratory visualisation,
it is worth taking the time to make sure that your main points are
easy to appreciate.

On [her excellent and helpful website](https://kateto.net/network-visualization), 
Katya Ognyanova outlines some key dimensions of control
that network researchers have to play with:

- vertex position (layout)
- vertex shape
- vertex color
- vertex size
- vertex labels
- edge color
- edge size
- edge shape
- edge arrows

## Approaches to visualising networks in R

There are a host of packages for plotting in R, 
and for plotting networks in R.
Plotting in R is typically based around two main approaches:
the 'base' approach in R by default,
and the 'grid' approach made popular by the famous and very flexible `{ggplot2}` package.^['gg' stands for the Grammar of Graphics.]
Approaches to plotting _graphs_ or _networks_ in R can be similarly divided.

The two classic packages are `{igraph}` and `{sna}`, both building upon the base R graphics engine.
Newer packages [`{ggnetwork}`](https://www.r-bloggers.com/2016/03/ggnetwork-network-geometries-for-ggplot2/) and 
[`{ggraph}`](https://ggraph.data-imaginist.com/index.html) build upon a grid approach.^[
Others include: 'Networkly' for creating 2-D and 3-D interactive networks
that can be rendered with plotly and can be easily integrated into
shiny apps or markdown documents;
'visNetwork' interacts with javascript (vis.js) to make interactive networks
(http://datastorm-open.github.io/visNetwork/); and
'networkD3' interacts with javascript (D3) to make interactive networks
(https://www.r-bloggers.com/2016/10/network-visualization-part-6-d3-and-r-networkd3/).
]
This vignette introduces some functions in the `{manynet}` package for plotting and 
visualising network data.
`{manynet}` builds upon the ggplot2/ggraph engine for plotting.

## Using `{manynet}` to quickly plot network graphs
  
To get a basic visualisation of the network before adding various specifications,
the `autographr()` function in `{manynet}` is a quick and easy way 
to obtain a clear first look of the network 
for preliminary investigations and understanding of the network.

```{r manyneteg, exercise=TRUE}

```
  
```{r manyneteg-solution}
library(manynet)
autographr(ison_brandes)
```
  
We can also specify the colours, groups, shapes, and sizes of nodes in 
the `autographr()` function using the following parameters:
  
* `node_colour`
* `node_shape`
* `node_size`

### Adding titles and subtitles

Append `ggtitle()` to add a title.
`{manynet}` works well with both `{ggplot2}` and `{ggraph}` functions 
that can be appended to create more tailored visualisations 
of the network.

```{r manynetexample2, exercise=TRUE}

```
  
```{r manynetexample2-solution}
autographr(ison_adolescents,
           labels = TRUE,
           node_size = 1.5) + 
  ggtitle("Visualisation")
```

`{manynet}` also uses the `{patchwork}` package for arranging
graphs together, e.g. side-by-side or above one another.
The syntax is quite straight forward and is used throughout these vignettes.

```{r patchwork, exercise=TRUE}

```

```{r patchwork-solution}
autographr(ison_adolescents) + autographr(ison_algebra)
autographr(ison_adolescents) / autographr(ison_algebra)
```

### Adding legends

By default, `{manynet}` doesn't add legends automatically, 
as often the colours are pretty self-explanatory,
for example in the following figure.

```{r maxbet}
ison_adolescents %>% 
  mutate(maxbet = node_is_max(node_betweenness(ison_adolescents))) %>% 
  autographr(node_color = "maxbet")
```

But other times it is important to add in a legend.
`{manynet}` supports the `{ggplot2}` way of adding legends
after the main plot has been constructed,
using `guides()` to add in the legends,
and `labs()` for giving those legends particular titles.

```{r discipline, exercise = TRUE}

```

```{r discipline-solution}
ison_networkers %>% 
  autographr(node_color = "Discipline") +
  guides(color = "legend") + 
  labs(color = "Discipline")
```

```{r discq}
question("What discipline is Charles Kadushin in according to this graph?",
         answer("Anthropology"),
         answer("Mathematics/Statistics"),
         answer("Other"),
         answer("Sociology", correct = TRUE),
        allow_retry = TRUE, random_answer_order = TRUE)
```

### Layouts

A range of graph layouts are available across the `{igraph}`, `{graphlayouts}`, and `{manynet}` packages 
that can be used together with `autographr()`.

```{r layoutseg, exercise=TRUE}

```

```{r layoutseg-solution}
(autographr(ison_southern_women, layout = "bipartite") + ggtitle("bipartite") |
autographr(ison_southern_women, layout = "hierarchy") + ggtitle("hierarchy")) /
(autographr(ison_southern_women, layout = "concentric") + ggtitle("concentric") |
   autographr(ison_southern_women, layout = "railway") + ggtitle("railway"))
```

### Using different colours

Because the autographr function is based on
the grammar of graphics,
it's easy to extend or alter aesthetic aspects.
Here let's try and change the colors
assigned to the different regions in the `mpn_elite_mex` dataset.

```{r colorch, exercise=TRUE}

```

```{r colorch-solution}
autographr(mpn_elite_mex,
           node_color = "region")

autographr(mpn_elite_mex,
           node_color = "region") +
  ggplot2::scale_colour_hue()

autographr(mpn_elite_mex,
           node_color = "region") +
  ggplot2::scale_colour_grey()

autographr(mpn_elite_mex,
           node_color = "region") +
  ggplot2::scale_colour_manual(
    values = c("1" = "red",
               "3" = "blue",
               "2" = "green"))
```

## Using `{ggraph}` for more flexibility

For more flexibility with visualizations, 
`{manynet}` users are encouraged to use the excellent `{ggraph}` package.
`{ggraph}` is built upon the venerable `{ggplot2}` package
and works with `tbl_graph` and `igraph` objects.
As with `{ggplot2}`, `{ggraph}` users are expected to build
a particular plot from the ground up,
adding explicit layers to visualise the nodes and edges.

```{r ggrapheg, exercise=TRUE}

```

```{r ggrapheg-solution}
library(ggraph)
ggraph(mpn_elite_mex, layout = "fr") + 
  geom_edge_link(edge_colour = "dark grey", 
                  arrow = arrow(angle = 45,
                                length = unit(2, "mm"),
                                type = "closed"),
                  end_cap = circle(3, "mm")) +
  geom_node_point(size = 2.5, shape = 19, colour = "blue") +
  geom_node_text(aes(label=name), family = "serif", size = 2.5) +
  scale_edge_width(range = c(0.3,1.5)) +
  theme_graph() +
  theme(legend.position = "none")
```
  
As we can see in the code above, we can specify various aspects of the plot to 
tailor it to our network. 
Firstly, we can alter the **layout** of the network using the `layout =` argument
to create a clearer visualisation of the ties between nodes. 
This is especially important for larger networks, where nodes and ties are more 
easily obscured or misrepresented.
In `{ggraph}`, the default layout is the "stress" layout. 
The "stress" layout is a safe choice because it is deterministic and 
fits well with almost any graph, but it is also a good idea to explore and try 
out other layouts on your data.
More layouts can be found in the `{graphlayouts}` and `{igraph}` R packages. 
To use a layout from the `{igraph}` package, enter only the last part of the layout 
algorithm name (eg. `layout = "mds"` for "layout_with_mds").
  
Secondly, using `geom_node_point()` which draws the nodes as geometric shapes 
(circles, squares, or triangles), we can specify the presentation of **nodes** 
in the network in terms of their *shape* (`shape=`, choose from 1 to 21), 
*size* (`size=`), or *colour* (`colour=`). We can also use `aes()` to match to 
node attributes. To add labels, use `geom_node_text()` or 
`geom_node_label()` (draws labels within a box). The font (`family=`), 
font size (`size=`), and colour (`colour=`) of the labels can be specified.
  
Thirdly, we can also specify the presentation of **edges** in the network.
To draw edges, we use `geom_edge_link0()` or `geom_edge_link()`. 
Using the latter function makes it possible to draw a straight line with a 
gradient. 
The following features can be tailored either globally or matched to specific
edge attributes using `aes()`:
  
* *colour*: `edge_colour=`
  
* *width*: `edge_width=`
  
* *linetype*: `edge_linetype=`
  
* *opacity*: `edge_alpha=`
  
For directed graphs, arrows can be drawn using the `arrow=` argument and the 
`arrow()` function from `{ggplot2}`. The angle, length, arrowhead type, and
padding between the arrowhead and the node can also be specified.
  
To change the position of the legend, add the `theme()` function from `{ggplot2}`.
The legend can be positioned at the top, bottom, left, or right, 
or removed using "none".

For more see David Schoch's [excellent resources on this](http://mr.schochastics.net/netVizR.html).
  
## Exporting plots to PDF

We can print the plots we have made to PDF by point-and-click
by selecting 'Save as PDF...' from under the 'Export' dropdown menu
in the plots panel tab of RStudio.

If you want to do this programmatically,
say because you want to record how you have saved it so that you can
e.g. make some changes to the parameters at some point,
this is also not too difficult.

After running the (gg-based) plot you want to save,
use the command `ggsave("my_filename.pdf")` to save your plot
as a PDF to your working directory.
If you want to save it somewhere else, you will need to specify the file path
(or change the working directory, but that might be more cumbersome).
If you want to save it as a different filetype,
replace `.pdf` with e.g. `.png` or `.jpeg`.
See `?ggsave` for more.
