---
title: "Topology"
author: "by James Hollway, Andrea Biswas-Tortajada"
output: 
  learnr::tutorial:
    theme: journal
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(manynet)
library(migraph)
library(patchwork)
knitr::opts_chunk$set(echo = FALSE)
```

## Generate networks of different structures

This tutorial covers a range of different network topologies:
trees, lattices, random, small-world, scale-free, and core-periphery
networks.
These ideal networks exaggerate centrality, cohesion, and randomness features, 
and are thus great for theory-building and investigating the relationship between rules and structure.

In this practical, we're going to create/generate
a number of ideal-typical network topologies and plot them.
We'll first look at some deterministic algorithms for _creating_ networks
of different structures, 
and then look at how the introduction of some randomness can _generate_ a variety of network structures.

## Deterministic graphs

To begin with, let's create a few 'empty' and full or 'complete' graphs.
You will want to use some of the `create_*()` group of functions from `{manynet}`,
because they create graphs following some strict rule(s).
The two functions you will want to use here are `create_empty()` and `create_filled()`.
`create_empty()` creates an empty graph with the given number of nodes, 
in this case 50 nodes.
For `create_filled()` we're creating a full graph,
where all of the nodes are connected to all of the other nodes.

Let's say that we want to explore networks of fifty nodes in this script.
Graph one empty and one complete network with 50 nodes each,
give them an informative title, and plot the graphs together.
What would a complete network with half the nodes look like?
Add that too.

```{r empty, exercise=TRUE}

```

```{r empty-solution}
(autographr(create_empty(50), "circle") + ggtitle("Empty graph"))
(autographr(to_undirected(create_filled(50))) + ggtitle("Complete graph"))
(autographr(to_undirected(create_filled(50/2))) + ggtitle("Complete graph (smaller)"))
```

### Stars

In a star network, there is one node to which all other nodes are connected.
There is no transitivity.
The maximum path length is two.
And centrality is maximised!
This network maximises all centrality measures as one node acts
as the sole bridge connecting one part of the network to the other.

Use the `create_star()` function to graph three star networks:

- an undirected star network
- a out-directed star network
- and an in-directed star network

```{r star, exercise = TRUE}

```

```{r star-solution}
(autographr(create_star(50)) + ggtitle("Star graph"))
(autographr(create_star(50, directed = TRUE)) + ggtitle("Star out"))
(autographr(to_redirected(create_star(50, directed = TRUE))) + ggtitle("Star in"))
```

### Trees

Trees, or regular trees, are networks with branching nodes.
They can be directed or undirected, and tend to indicate strong hierarchy.
Again graph three networks:

- one undirected with 2 branches per node
- a directed network with 2 branches per node
- the same as above, but graphed using the "tree" layout

```{r tree, exercise = TRUE}

```

```{r tree-solution}
(autographr(create_tree(50, width = 2)) + ggtitle("Tree graph"))
(autographr(create_tree(50, width = 2, directed = TRUE)) + ggtitle("Tree out"))
(autographr(create_tree(50, width = 2, directed = TRUE), "tree") + ggtitle("Tree layout"))
```

Try varying the `width` argument to see the result.

### Lattices

Lattices reflect highly clustered networks
where there is a high likelihood that interaction partners also interact.
They are used to show how clustering facilitates or limits diffusion
or makes pockets of behaviour stable.

Note that `create_lattice()` in `{manynet}` works a little differently
to how it works in `{igraph}`.
In `{igraph}` the number or vector passed to the function indicates
the length of each dimension.
So `c(50)` would be a one-dimensional lattice,
essentially a chain of 50 nodes connected to their neighbours.
`c(50,50)` would be a two-dimensional lattice,
of 50 nodes long and 50 nodes wide.
`c(50,50,50)` would be a three-dimensional lattice,
of 50 nodes long, 50 nodes wide, and 50 nodes deep, etc.

_But_ this doesn't help us when we want to see what a lattice representation
with the same order (number of nodes) as a given network would be.
For example, perhaps we just want to know what a lattice with 50 nodes
would look like.
So `{manynet}` instead tries to find the most even or balanced 
two-dimensional representation with a given number of nodes.

Graph two lattices, one with 50 nodes, 
and another with half the number of nodes.

```{r lattices, exercise = TRUE}

```

```{r lattices-solution}
(autographr(create_lattice(50)) + ggtitle("One-mode lattice graph"))
(autographr(create_lattice(50/2)) + ggtitle("Smaller lattice graph"))
```

### Rings

This creates a graph where each node has two separate neighbours
which creates a ring graph.
Graph three ring networks:

- one with 50 nodes
- one with 50 nodes where they are connected to neighbours two steps away,
  on a "circle" layout
- the same as above, but on a "stress" layout

```{r rings, exercise = TRUE}

```

```{r rings-solution}
(autographr(create_ring(50)) + ggtitle("Ring graph", subtitle = "Starring Naomi Watts"))
(autographr(create_ring(50, width = 2), "circle") + ggtitle("The Ring Two", subtitle = "No different?"))
(autographr(create_ring(50, width = 2), "stress") + ggtitle("The Ring Two v2.0"))
```

## Probabilistic graphs

Next we are going to take a look at some probabilistic graphs.
These involve some random element, perhaps in addition to specific rules,
to stochastically 'generate' networks of certain types of topologies.
As such, we'll be using the `generate_*()` group of functions from `{manynet}`.

### Random graphs

An Erdös-Renyi graph is simply a random graph.
You will need to specify the probability of a tie
in addition to the number of nodes.
An An Erdos-Renyi  graph on the vertex set $V$ is a random graph
which connects each pair of nodes ${i,j}$ with probability $p$, independent.
Note that for a “sparse” ER graphs, $p$ must decrease as $N$ goes up.
Generate three random networks of 50 nodes and a density of 0.08:

```{r random, exercise = TRUE}

```

```{r random-solution}
(autographr(generate_random(50, 0.08)) + ggtitle("Random 1 graph"))
(autographr(generate_random(50, 0.08)) + ggtitle("Random 2 graph"))
(autographr(generate_random(50, 0.08)) + ggtitle("Random 3 graph"))
```

Keep going if you like... it will be a little different every time.
Note that you can also pass the second argument an integer,
in which case the function will interpret that as the number of ties/edges rather than the probability that a tie is present.
Try generating a random graph with 200 edges/ties now:

```{r randomno, exercise = TRUE}

```

```{r randomno-solution}
(erdren4 <- autographr(generate_random(50, 200)) + ggtitle("Random 1 graph"))
```

### Small-world graphs

Remember the ring graph from above?
What if we rewire (change) some of the edges at a certain probability?
This is how small-world networks are generated.
Graph three small-world networks, all with 50 nodes and a rewiring probability of 0.025.

```{r smallw, exercise = TRUE}

```

```{r smallw-solution}
(autographr(generate_smallworld(50, 0.025)) + ggtitle("Smallworld 1 graph"))
(autographr(generate_smallworld(50, 0.025)) + ggtitle("Smallworld 2 graph"))
(autographr(generate_smallworld(50, 0.025)) + ggtitle("Smallworld 3 graph"))
```

With on average 2.5 ties randomly rewired, does the structure look different?
This is a small-world network, where clustering/transitivity remains high
but path lengths are much lower than they would otherwise be.
Remember that in a small-world network, the shortest-path  distance between nodes
increases sufficiently slowly as a function of the number of nodes in the network.
You can also call these networks a Watts–Strogatz toy network.
If you want to review this, go back to the reading by Watts (2004).

There is also such a thing as a network's small-world coefficient.
See the help page for more details,
but with the default equation ('omega'),
the coefficient typically ranges between 0 and 1,
where 1 is as close to a small-world as possible.
Try it now on a small-world generated network,
but with a rewiring probability of 0.25:

```{r smallwtest, exercise = TRUE}

```

```{r smallwtest-solution}
network_smallworld(generate_smallworld(50, 0.25))
```

### Scale-free graphs

There is another famous model in network science: the scale-free model.
Remember:
"In many real-world networks, the distribution of the number of network neighbours
the degree distribution is typically right-skewed with a "heavy tail".
A majority of the nodes have less-than-average degree and
a small fraction of hubs are many times better connected than average (2004, p. 250).

The following generates a scale-free graph according to the Barabasi-Albert (BA) model
that rests upon the mechanism of preferential attachment.
More on this on the Watts paper (2005, p.51) and Merton (1968)
The BA model rests on two mechanisms: 
population growth and preferential attachment.
Population growth: real networks grow in time as new members join the population.
Preferential/cumulative attachment means that newly arriving nodes will tend to
connect to already well-connected nodes rather than poorly connected ones.

Generate and graph three scale-free networks,
with alpha parameters of 0.5, 1, and 1.5.

```{r scalef, exercise = TRUE}

```

```{r scalef-solution}
(autographr(generate_scalefree(50, 0.5)) +
    ggtitle("Scalefree 1 graph", subtitle = "Power = .5"))
(autographr(generate_scalefree(50, 1)) +
    ggtitle("Scalefree 2 graph", subtitle = "Power = 1"))
(autographr(generate_scalefree(50, 1.5)) +
    ggtitle("Scalefree 3 graph", subtitle = "Power = 1.5"))
```

You can also test whether a network has a degree distribution that fits
the scale-free model.
When a Kolmogorov-Smirnov test p-value less than 0.05 is implied,
a message is given that you should reject the hypothesis 
that a power law fits here.
With an alpha/power-law exponent between 2 and 3, 
one generally cannot reject the hypothesis that the observed data 
comes from a power-law distribution.

```{r scaleftest, exercise = TRUE}

```

```{r scaleftest-solution}
network_scalefree(generate_scalefree(50, 2))
```

### Core-periphery graphs

Lastly, we'll take a look at some core-periphery graphs.
The most common definition of a core-periphery network
is one in which the network can be partitioned into two groups
such that one group of nodes (the core) has 
dense interactions among themselves,
moderately dense interactions with the second group,
and the second group (the periphery) has
verse sparse interactions among themselves.

We can visualise extreme versions of such a network
using the `create_core()` function.
Graph a core-periphery network of 50 nodes
(which, unless a core-periphery membership assignment is given,
will be split evenly between core and periphery partitions).

```{r core, exercise=TRUE}

```

```{r core-solution}
(autographr(create_core(50)) + ggtitle("Core"))
```

Let's consider identifying the core and peripheral nodes
in a network.
Let's use the `ison_brandes` dataset from `{manynet}`,
and identify the core/periphery assignment of the _nodes_.
Graph the data with the core/periphery assignment.

```{r nodecore, exercise=TRUE}

```

```{r nodecore-solution}
ison_brandes %>% 
  mutate(nc = node_core(ison_brandes)) %>% 
  autographr(node_color = "nc")
```

An alternative route is to identify 'core' nodes
depending on their _k_-coreness.
In `{migraph}`, we can return nodes _k_-coreness
with `node_coreness()` instead of 
the `node_core()` used for core-periphery.

```{r nodecoren, exercise=TRUE}

```

```{r nodecoren-solution}
ison_brandes %>% 
  mutate(ncn = node_coreness(ison_brandes)) %>% 
  autographr(node_color = "ncn")
```
