---
title: "Diversity and Regression"
author: "by James Hollway"
output: 
  learnr::tutorial:
    theme: journal
runtime: shiny_prerendered
description: >
  This tutorial aims to teach you how to measure and test network diversity,
  moving from univariate to multivariate tests, including 
  network linear models (multiple regression quadratic assignment procedures).
---

```{r setup, include=FALSE}
library(learnr)
library(manynet)
library(migraph)
library(patchwork)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
marvel_friends <- to_unsigned(ison_marvel_relationships, keep = "positive")
marvel_friends <- to_giant(marvel_friends)
marvel_friends <- marvel_friends %>% to_subgraph(Appearances >= mean(Appearances))
```

## Initial visualisation

For this session, we'll explore a couple of different datasets.
First, let's examine homogeneity/heterogeneity in the Marvel relationships dataset from `{manynet}`,
`ison_marvel_relationships`.
The dataset is quite complicated,
so to make this simpler, let's concentrate on:

- just the positive (friendship) ties and not the negative (enmity) ties
- the main (giant) component without any isolates
- just those characters that appear in the comics more than average

Fortunately, all these data cleaning moves are easy to do in `{manynet}`,
and can be seen in the following chunk in order:

```{r friends, exercise=TRUE, purl = FALSE}

```

```{r friends-hint-1, purl = FALSE}
# since the dataset is a 'signed' graph, we want to get just the
# positively signed ties to get the friendship graph 
# (and lose the enmity relations)
to_unsigned(____, keep = "positive")
```

```{r friends-hint-2, purl = FALSE}
# to_giant() is a quick easy way to get the giant/main component
to_giant(____)
```

```{r friends-hint-3, purl = FALSE}
to_subgraph(____, Appearances >= mean(Appearances))
```

```{r friends-hint-4, purl = FALSE}
# don't forget to assign the results!
marvel_friends <- ____
```

```{r friends-hint-5, purl = FALSE}
marvel_friends <- to_unsigned(ison_marvel_relationships, keep = "positive")
marvel_friends <- to_giant(marvel_friends)
marvel_friends <- marvel_friends %>% to_subgraph(Appearances >= mean(Appearances))
marvel_friends
```

```{r friends-solution}
marvel_friends <- to_unsigned(ison_marvel_relationships, keep = "positive")
marvel_friends <- to_giant(marvel_friends)
marvel_friends <- marvel_friends %>% to_subgraph(Appearances >= mean(Appearances))
marvel_friends
```

This gives us an undirected network of nearly twenty characters and a little more than 100 edges.
Recall that this data has several nodal attributes.
Let's explore a couple of these attributes, "Gender" and "PowerOrigin", visually
using `graphr()`.

```{r plotfriends, exercise=TRUE, purl = FALSE}

```

```{r plotfriends-hint, purl = FALSE}
# Since both Gender and PowerOrigin are categorical variables
# you will need to use two different aesthetic dimensions to
# represent them together.
# Which will you present as shapes and which as colors?
graphr(____, 
       node_shape = ____,
       node_color = ____)
```

```{r plotfriends-solution}
graphr(marvel_friends, 
           node_shape = "Gender",
           node_color = "PowerOrigin")
```

These variables seem to be distributed unevenly across the network.
There seems to be some homophily -- or like choosing like --
operating here, but it is difficult to say conclusively because there are
clearly more male than female superheros,
as well as clearly more superheros of mutant origin than others.
So what might seem like homophily could just be a result of there being
many more opportunities for ties between nodes of some categories.
We therefore need to establish how diverse this network really is.

